---
title: 'Assessment Part 1: Properties of Stars'
author: "kevinoliva26"
date: "12/1/2020"
output: html_document
---

Load the stars data frame from dslabs. This contains the name, absolute magnitude, temperature in degrees Kelvin, and spectral class of selected stars. Absolute magnitude (shortened in these problems to simply "magnitude") is a function of star luminosity, where negative values of magnitude have higher luminosity.
```{r}
library(tidyverse)
library(dslabs)
data(stars)
options(digits = 3) 

#What is the mean magnitude?
mean(stars$magnitude)
#What is the standard deviation of magnitude?
sd(stars$magnitude)

#Make a density plot of the magnitude.
#How many peaks are there in the data?
stars %>% 
  ggplot(aes(magnitude)) +
  geom_density()

#Examine the distribution of star temperature.
#Which of these statements best characterizes the temperature distribution?
stars %>%
  ggplot(aes(temp)) +
  geom_density()

#Make a scatter plot of the data with temperature on the x-axis and magnitude on the y-axis and examine the relationship between the variables. Recall that lower magnitude means a more luminous (brighter) star.
#When considering the plot of magnitude vs. temperature, most stars follow a _______________ trend. These are called main sequence stars.
stars %>%
  ggplot(aes(temp,magnitude)) +
  geom_point()

#For various reasons, scientists do not always follow straight conventions when making plots, and astronomers usually transform values of star luminosity and temperature before plotting. Flip the y-axis so that lower values of magnitude are at the top of the axis (recall that more luminous stars have lower magnitude) using scale_y_reverse(). Take the log base 10 of temperature and then also flip the x-axis.
#The brighest, highest temperature stars are in the ______________ corner of the plot.
stars %>%
  ggplot(aes(temp, magnitude, color = type)) +
  geom_point() +
  scale_x_continuous(trans = "log10") +
  scale_y_continuous(trans = "log10") +
  scale_y_reverse() +
  scale_x_reverse() +
  geom_vline(xintercept = 5000)

#The trends you see allow scientists to learn about the evolution and lifetime of stars. The primary group of stars to which most stars belong we will call the main sequence stars (discussed in question 4). Most stars belong to this main sequence, however some of the more rare stars are classified as “old” and “evolved” stars. These stars tend to be hotter stars, but also have low luminosity, and are known as white dwarfs.
#How many white dwarfs are there in our sample?
stars %>%
  ggplot(aes(temp, magnitude, color = type)) +
  geom_point() +
  scale_x_continuous(trans = "log10") +
  scale_y_continuous(trans = "log10") +
  scale_y_reverse() +
  scale_x_reverse() +
  geom_vline(xintercept = 5000)

#Consider stars which are not part of the Main Group but are not old/evolved (white dwarf) stars. These stars must also be unique in certain ways and are known as giants. Use the plot from Question 5 to estimate the average temperature of a giant.

#Which of these temperatures is closest to the average temperature of a giant?:
stars %>%
  ggplot(aes(temp, magnitude, color = type)) +
  geom_point() +
  scale_x_continuous(trans = "log10") +
  scale_y_continuous(trans = "log10") +
  scale_y_reverse() +
  scale_x_reverse() +
  geom_vline(xintercept = 5000)

#We can now identify whether specific stars are main sequence stars, red giants or white dwarfs. Add text labels to the plot to answer these questions. You may wish to plot only a selection of the labels, repel the labels, or zoom in on the plot in RStudio so you can locate specific stars.
library(ggrepel)
stars %>%
  ggplot(aes(temp, magnitude, color = type, label = star)) +
  geom_point() +
  scale_x_continuous(trans = "log10") +
  scale_y_continuous(trans = "log10") +
  scale_y_reverse() +
  scale_x_reverse() +
  geom_text(size = 3) +
  geom_vline(xintercept = 5000) +
  scale_colour_brewer(palette="Set1")

stars %>%
  filter(temp < 5000) %>%
  ggplot(aes(temp, magnitude, color = type, label = star)) +
  geom_point() +
  scale_x_continuous(trans = "log10") +
  scale_y_continuous(trans = "log10") +
  scale_y_reverse() +
  scale_x_reverse() +
  geom_text(size = 3) +
  geom_vline(xintercept = 5000) 
```

Load the temp_carbon dataset from dslabs, which contains annual global temperature anomalies (difference from 20th century mean temperature in degrees Celsius), temperature anomalies over the land and ocean, and global carbon emissions (in metric tons). Note that the date ranges differ for temperature and carbon emissions.
```{r}
#Which of these code blocks return the latest year for which carbon emissions are reported?
library(tidyverse)
library(dslabs)
data(temp_carbon)
data(greenhouse_gases)
data(historic_co2)

temp_carbon %>%
  filter(!is.na(carbon_emissions)) %>%
  pull(year) %>%
  max()

temp_carbon %>%
  filter(!is.na(carbon_emissions)) %>%
  .$year %>%
  max()

temp_carbon %>%
  filter(!is.na(carbon_emissions)) %>%
  select(year) %>%
  max()

#What is the first year for which carbon emissions (carbon_emissions) data are available?
temp_carbon %>%
  group_by(year) %>%
  head()

#What is the first year for which carbon emissions (carbon_emissions) data are available?
temp_carbon %>%
  filter(!is.na(carbon_emissions)) %>%
  pull(year) %>%
  min()

#How many times larger were carbon emissions in the last year relative to the first year?
last_year <- temp_carbon %>%
  filter(year== 2014) %>%
  pull(carbon_emissions) 

first_year <- temp_carbon %>%
  filter(year== 1751) %>%
  pull(carbon_emissions) 

last_year/first_year

#Inspect the difference in temperature in temp_carbon from the first available year to the last available year.
#What is the first year for which global temperature anomaly (temp_anomaly) data are available?
temp_carbon %>%
  filter(!is.na(temp_anomaly)) %>%
  pull(year) %>%
  min()

#What is the last year for which global temperature anomaly data are available?
temp_carbon %>%
  filter(!is.na(temp_anomaly)) %>%
  pull(year) %>%
  max()

#How many degrees Celsius has temperature increased over the date range? Compare the temperatures in the most recent year versus the oldest year.
temp1 <- temp_carbon %>%
  filter(!is.na(temp_anomaly) & year == 1880) %>%
  pull(temp_anomaly)

temp2 <- temp_carbon %>%
  filter(!is.na(temp_anomaly) & year == 2018) %>%
  pull(temp_anomaly)

temp2-temp1

#Create a time series line plot of the temperature anomaly. Only include years where temperatures are reported. Save this plot to the object p.
#Which command adds a blue horizontal line indicating the 20th century mean temperature?
p <- temp_carbon %>%
  filter(!is.na(temp_anomaly)) %>%
  ggplot(aes(year,temp_anomaly)) +
  geom_line()
p <- p + geom_hline(aes(yintercept = 0), col = "blue")
p

#Continue working with p, the plot created in the previous question.
#Change the y-axis label to be "Temperature anomaly (degrees C)". Add a title, "Temperature anomaly relative to 20th century mean, 1880-2018". Also add a text layer to the plot: the x-coordinate should be 2000, the y-coordinate should be 0.05, the text should be "20th century mean", and the text color should be blue.
#Which of the following code blocks is correct?
p <- p + ylab("Temperature anomaly (degrees C)") +
  ggtitle("Temperature anomaly relative to 20th century mean, 1880-2018") +
  geom_text(aes(x = 2000, y = 0.05, label = "20th century mean"), col = "blue")

#Add layers to the previous plot to include line graphs of the temperature anomaly in the ocean (ocean_anomaly) and on land (land_anomaly). Assign different colors to the lines. Compare the global temperature anomaly to the land temperature anomaly and ocean temperature anomaly.
p + geom_line(aes(year, land_anomaly), col = "red") +
  geom_line(aes(year,ocean_anomaly), col="brown")
```

The climate change exercises continue on this page. Use the libraries and datasets below.
```{r}
library(tidyverse)
library(dslabs)
data(temp_carbon)
data(greenhouse_gases)
data(historic_co2)

#Complete the code outline below to make a line plot of concentration on the y-axis by year on the x-axis. Facet by gas, aligning the plots vertically so as to ease comparisons along the year axis. Add a vertical line with an x-intercept at the year 1850, noting the unofficial start of the industrial revolution and widespread fossil fuel consumption. Note that the units for ch4 and n2o are ppb while the units for co2 are ppm.
greenhouse_gases %>%
    ggplot(aes(year,concentration)) +
    geom_line() +
    facet_grid(gas ~ ., scales = "free") +
    geom_vline(xintercept = 1850) +
    ylab("Concentration (ch4/n2o ppb, co2 ppm)") +
    ggtitle("Atmospheric greenhouse gas concentration by year, 0-2000")

#Make a time series line plot of carbon emissions (carbon_emissions) from the temp_carbon dataset. The y-axis is metric tons of carbon emitted per year.
temp_carbon %>%
  filter(!is.na(carbon_emissions)) %>%
  ggplot(aes(year,carbon_emissions)) +
  geom_line() +
  geom_vline(xintercept = 1970) +
  geom_vline(xintercept = 2014)

#Make a line plot of co2 concentration over time (year), coloring by the measurement source (source). Save this plot as co2_time for later use.
#Which of the following are true about co2_time, the time series of co2 over the last 800,000 years?
co2_time <- historic_co2 %>%
  filter(!is.na(co2) & !is.na(year)) %>%
  ggplot(aes(year,co2)) +
  geom_line()

#Use the co2_time plot saved above. Change the limits as directed to investigate the rate of change in co2 over various periods with spikes in co2 concentration.
#Change the x-axis limits to -800,000 and -775,000. About how many years did it take for co2 to rise from 200 ppmv to its peak near 275 ppmv?
co2_time +
  xlim(c(-800000,-775000))

#Change the x-axis limits to -375,000 and -330,000. About how many years did it take for co2 to rise from the minimum of 180 ppm to its peak of 300 ppmv?
co2_time +
  xlim(c(-375000,-330000))

#Change the x-axis limits to -140,000 and -120,000. About how many years did it take for co2 to rise from 200 ppmv to its peak near 280 ppmv?
co2_time +
  xlim(c(-140000,-120000))

#Change the x-axis limits to -3000 and 2018 to investigate modern changes in co2. About how many years did it take for co2 to rise from its stable level around 275 ppmv to the current level of over 400 ppmv?
co2_time +
  xlim(c(-3000,2018))
```




