---
title: 'Section 5: Data Visualization Principles'
author: "kevinoliva26"
date: "11/27/2020"
output: html_document
---

##Encoding Data Using Visual Cues

Visual cues for encoding data include position, length, angle, area, brightness and color hue.
Position and length are the preferred way to display quantities, followed by angles, which are preferred over area. Brightness and color are even harder to quantify but can sometimes be useful.
Pie charts represent visual cues as both angles and area, while donut charts use only area. Humans are not good at visually quantifying angles and are even worse at quantifying area. Therefore pie and donut charts should be avoided - use a bar plot instead. If you must make a pie chart, include percentages as labels.
Bar plots represent visual cues as position and length. Humans are good at visually quantifying linear measures, making bar plots a strong alternative to pie or donut charts.

##Know When to Include Zero

When using bar plots, always start at 0. It is deceptive not to start at 0 because bar plots imply length is proportional to the quantity displayed. Cutting off the y-axis can make differences look bigger than they actually are.

When using position rather than length, it is not necessary to include 0 (scatterplot, dot plot, boxplot).

##Do Not Distort Quantities

Make sure your visualizations encode the correct quantities.
For example, if you are using a plot that relies on circle area, make sure the area (rather than the radius) is proportional to the quantity.

##Order by a Meaningful Value

It is easiest to visually extract information from a plot when categories are ordered by a meaningful value. The exact value on which to order will depend on your data and the message you wish to convey with your plot.
The default ordering for categories is alphabetical if the categories are strings or by factor level if factors. However, we rarely want alphabetical order.

##Show the Data

A dynamite plot - a bar graph of group averages with error bars denoting standard errors - provides almost no information about a distribution.

By showing the data, you provide viewers extra information about distributions.
Jitter is adding a small random shift to each point in order to minimize the number of overlapping points. To add jitter, use the  *geom_jitter()* geometry instead of *geom_point()*. 
*Alpha* blending is making points somewhat transparent, helping visualize the density of overlapping points. Add an alpha argument to the geometry.

```{r}
# dot plot showing the data
heights %>% ggplot(aes(sex, height)) + geom_point()

# jittered, alpha blended point plot
heights %>% ggplot(aes(sex, height)) + geom_jitter(width = 0.1, alpha = 0.2)
```

##Ease Comparisons: Use Common Axes

Ease comparisons by keeping axes the same when comparing data across multiple plots.
Align plots vertically to see horizontal changes. Align plots horizontally to see vertical changes.
Bar plots are useful for showing one number but not useful for showing distributions.

##Consider Transformations

Use transformations when warranted to ease visual interpretation.
The log transformation is useful for data with multiplicative changes. The logistic transformation is useful for fold changes in odds. The square root transformation is useful for count data.

##Ease Comparisons: Compared Visual Cues Should Be Adjacent

When two groups are to be compared, it is optimal to place them adjacent in the plot.
Use color to encode groups to be compared.
Consider using a color blind friendly palette like the one in this video.

```{r}
color_blind_friendly_cols <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

p1 <- data.frame(x = 1:8, y = 1:8, col = as.character(1:8)) %>%
    ggplot(aes(x, y, color = col)) +
    geom_point(size = 5)
p1 + scale_color_manual(values = color_blind_friendly_cols)
```

##Exercises

Using the reorder funtion inside a mutate
```{r}
library(dplyr)
library(ggplot2)
library(dslabs)
dat <- us_contagious_diseases %>% filter(year == 1967 & disease=="Measles" & count>0 & !is.na(population)) %>%
 mutate(rate = count / population * 10000 * 52 / weeks_reporting) %>%
 mutate(state = reorder(state, rate))
dat %>% ggplot(aes(state, rate)) +
 geom_bar(stat="identity") +
 coord_flip()
```

Reorder by median
```{r}
library(dplyr)
library(ggplot2)
library(dslabs)
data("murders")
murders %>% mutate(rate = total/population*100000) %>% mutate(region = reorder(region, rate,FUN = median)) %>% ggplot(aes(region,rate)) + geom_boxplot() + geom_point()
```

##Slope Charts

Consider using a slope chart or Bland-Altman plot when comparing one variable at two different time points, especially for a small number of observations.
Slope charts use angle to encode change. Use *geom_line()* to create slope charts. It is useful when comparing a small number of observations.
The *Bland-Altman plot* (Tukey mean difference plot, MA plot) graphs the difference between conditions on the y-axis and the mean between conditions on the x-axis. It is more appropriate for large numbers of observations than slope charts.

Slope Chart
```{r}
library(tidyverse)
library(dslabs)
data(gapminder)

west <- c("Western Europe", "Northern Europe", "Southern Europe", "Northern America", "Australia and New Zealand")

dat <- gapminder %>%
    filter(year %in% c(2010, 2015) & region %in% west & !is.na(life_expectancy) & population > 10^7)

dat %>%
    mutate(location = ifelse(year == 2010, 1, 2),
           location = ifelse(year == 2015 & country %in% c("United Kingdom", "Portugal"),
                             location + 0.22, location),
           hjust = ifelse(year == 2010, 1, 0)) %>%
    mutate(year = as.factor(year)) %>%
    ggplot(aes(year, life_expectancy, group = country)) +
    geom_line(aes(color = country), show.legend = FALSE) +
    geom_text(aes(x = location, label = country, hjust = hjust), show.legend = FALSE) +
    xlab("") +
    ylab("Life Expectancy") 
```

Bland-Altman plot
```{r}
library(ggrepel)
dat %>%
    mutate(year = paste0("life_expectancy_", year)) %>%
    select(country, year, life_expectancy) %>% spread(year, life_expectancy) %>%
    mutate(average = (life_expectancy_2015 + life_expectancy_2010)/2,
                difference = life_expectancy_2015 - life_expectancy_2010) %>%
    ggplot(aes(average, difference, label = country)) +
    geom_point() +
    geom_text_repel() +
    geom_abline(lty = 2) +
    xlab("Average of 2010 and 2015") +
    ylab("Difference between 2015 and 2010")
```

##Encoding a Third Variable

Encode a categorical third variable on a scatterplot using color hue or shape. Use the *shape* argument to control shape.
Encode a continuous third variable on a using *color intensity* or *size.*

Vaccines save millions of lives, but misinformation has led some to question the safety of vaccines. The data support vaccines as safe and effective. We visualize data about measles incidence in order to demonstrate the impact of vaccination programs on disease rate.
The *RColorBrewer* package offers several color palettes. Sequential color palettes are best suited for data that span from high to low. Diverging color palettes are best suited for data that are centered and diverge towards high or low values.
The *geom_tile()* geometry creates a grid of colored tiles.
Position and length are stronger cues than color for numeric values, but color can be appropriate sometimes.

Tile plot of measles rate by year and state
```{r}
# import data and inspect
library(tidyverse)
library(dslabs)
data(us_contagious_diseases)
str(us_contagious_diseases)

# assign dat to the per 10,000 rate of measles, removing Alaska and Hawaii and adjusting for weeks reporting
the_disease <- "Measles"
dat <- us_contagious_diseases %>%
    filter(!state %in% c("Hawaii", "Alaska") & disease == the_disease) %>%
    mutate(rate = count / population * 10000 * 52/weeks_reporting) %>%
    mutate(state = reorder(state, rate))

# plot disease rates per year in California
dat %>% filter(state == "California" & !is.na(rate)) %>%
    ggplot(aes(year, rate)) +
    geom_line() +
    ylab("Cases per 10,000") +
    geom_vline(xintercept=1963, col = "blue")

# tile plot of disease rate by state and year
dat %>% ggplot(aes(year, state, fill=rate)) +
    geom_tile(color = "grey50") +
    scale_x_continuous(expand = c(0,0)) +
    scale_fill_gradientn(colors = RColorBrewer::brewer.pal(9, "Reds"), trans = "sqrt") +
    geom_vline(xintercept = 1963, col = "blue") +
    theme_minimal() + theme(panel.grid = element_blank()) +
    ggtitle(the_disease) +
    ylab("") +
    xlab("")
```

Line plot of measles rate by year and state
```{r}
# compute US average measles rate by year
avg <- us_contagious_diseases %>%
    filter(disease == the_disease) %>% group_by(year) %>%
    summarize(us_rate = sum(count, na.rm = TRUE)/sum(population, na.rm = TRUE)*10000)

# make line plot of measles rate by year by state
dat %>%
    filter(!is.na(rate)) %>%
    ggplot() +
    geom_line(aes(year, rate, group = state), color = "grey50", 
        show.legend = FALSE, alpha = 0.2, size = 1) +
    geom_line(mapping = aes(year, us_rate), data = avg, size = 1, col = "black") +
    scale_y_continuous(trans = "sqrt", breaks = c(5, 25, 125, 300)) +
    ggtitle("Cases per 10,000 by state") +
    xlab("") +
    ylab("") +
    geom_text(data = data.frame(x = 1955, y = 50),
        mapping = aes(x, y, label = "US average"), color = "black") +
    geom_vline(xintercept = 1963, col = "blue")
```

##Avoid Pseudo and Gratuitous 3D Plots

In general, pseudo-3D plots and gratuitous 3D plots only add confusion. Use regular 2D plots instead.

##Avoid Too Many Significant Digits

In tables, avoid using too many significant digits. Too many digits can distract from the meaning of your data.
Reduce the number of significant digits globally by setting an option. For example, *options(digits = 3)* will cause all future computations that session to have 3 significant digits.
Reduce the number of digits locally using *round()* or *signif()*.

##Data Camp example
```{r}
library(dplyr)
library(ggplot2)
library(dslabs)
library(RColorBrewer)
data(us_contagious_diseases)

us_contagious_diseases %>% filter(!is.na(population)) %>% 
  group_by(year, disease) %>%
  summarize(rate = sum(count)/sum(population)*10000) %>%
  ggplot(aes(year, rate, color = disease)) + 
  geom_line()
```

##Assessment: Titanic Survival

Define the titanic dataset starting from the titanic library with the following code:
```{r}
options(digits = 3)    # report 3 significant digits
library(tidyverse)
library(titanic)

titanic <- titanic_train %>%
    select(Survived, Pclass, Sex, Age, SibSp, Parch, Fare) %>%
    mutate(Survived = factor(Survived),
           Pclass = factor(Pclass),
           Sex = factor(Sex))
```

Make density plots of age grouped by sex. Try experimenting with combinations of faceting, alpha blending, stacking and using variable counts on the y-axis to answer the following questions. Some questions may be easier to answer with different versions of the density plot.
```{r}
library(dplyr)
titanic %>% ggplot(aes(Age, color=Sex)) + geom_density(alpha=0.5) + geom_vline(xintercept = 5, col = "blue") + geom_vline(xintercept = 25, col = "red") 

titanic %>% filter(Age >=40) %>% ggplot(aes(Sex)) + geom_bar()

titanic %>% filter(Age >=18 & Age <=35) %>% ggplot(aes(Sex)) + geom_bar()

titanic %>% filter(Age < 17) %>% ggplot(aes(Sex)) + geom_bar()

#Oldest individual
titanic[which.max(titanic$Age),"Sex"]
```

Use geom_qq() to make a QQ-plot of passenger age and add an identity line with geom_abline(). Filter out any individuals with an age of NA first. Use the following object as the dparams argument in geom_qq():
```{r}
library(titanic)
titanic <- titanic_train %>%
    select(Survived, Pclass, Sex, Age, SibSp, Parch, Fare) %>%
    mutate(Survived = factor(Survived),
           Pclass = factor(Pclass),
           Sex = factor(Sex))

titanic %>%
    filter(!is.na(Age)) %>%
    ggplot(aes(sample = Age)) +
    geom_qq(dparams = params) +
    geom_abline()
```

To answer the following questions, make barplots of the Survived and Sex variables using geom_bar(). Try plotting one variable and filling by the other variable. You may want to try the default plot, then try adding position = position_dodge() to geom_bar() to make separate bars for each group.
```{r}
#plot 1 - survival filled by sex
titanic %>%
    ggplot(aes(Survived, fill = Sex)) +
    geom_bar()
# plot 2 - survival filled by sex with position_dodge
titanic %>%
    ggplot(aes(Survived, fill = Sex)) +
    geom_bar(position = position_dodge())
#plot 3 - sex filled by survival
titanic %>%
    ggplot(aes(Sex, fill = Survived)) +
    geom_bar()
```

Make a density plot of age filled by survival status. Change the y-axis to count and set alpha = 0.2.

Which age group is the only group more likely to survive than die?
```{r}
#Which age group is the only group more likely to survive than die?
titanic %>%
    ggplot(aes(Age, fill=Survived)) +
    geom_density(alpha = 0.2) +
    geom_vline(xintercept = 70) +
    geom_vline(xintercept = 80)

#Which age group had the most deaths?
#Which age group had the highest proportion of deaths?
```


Filter the data to remove individuals who paid a fare of 0. Make a boxplot of fare grouped by survival status. Try a log2 transformation of fares. Add the data points with jitter and alpha blending.

```{r}
titanic %>%
    filter(Fare != 0) %>%
    ggplot(aes(Fare, Survived)) +
    geom_boxplot() +
    scale_x_continuous(trans = "log2") +
    geom_jitter(width = 0.1, alpha = 0.2)
```

The Pclass variable corresponds to the passenger class. Make three barplots. For the first, make a basic barplot of passenger class filled by survival. For the second, make the same barplot but use the argument position = position_fill() to show relative proportions in each group instead of counts. For the third, make a barplot of survival filled by passenger class using position = position_fill().
```{r}
#barplot 1
titanic %>%
    ggplot(aes(Pclass, fill = Survived)) +
    geom_bar()

#barplot 2
titanic %>%
    ggplot(aes(Pclass, fill = Survived)) +
    geom_bar(position = position_fill())

#barplot 3
titanic %>%
    ggplot(aes(Survived, fill = Pclass)) +
    geom_bar(position = position_fill())
```

Create a grid of density plots for age, filled by survival status, with count on the y-axis, faceted by sex and passenger class.
```{r}
titanic %>%
    ggplot(aes(Age, ..count..,fill = Survived)) +
    geom_density(alpha = 0.2) +
    facet_grid(Sex ~ Pclass)
```

