---
title: 'Section 5: Data Visualization Principles'
author: "kevinoliva26"
date: "11/27/2020"
output: html_document
---

##Encoding Data Using Visual Cues

Visual cues for encoding data include position, length, angle, area, brightness and color hue.
Position and length are the preferred way to display quantities, followed by angles, which are preferred over area. Brightness and color are even harder to quantify but can sometimes be useful.
Pie charts represent visual cues as both angles and area, while donut charts use only area. Humans are not good at visually quantifying angles and are even worse at quantifying area. Therefore pie and donut charts should be avoided - use a bar plot instead. If you must make a pie chart, include percentages as labels.
Bar plots represent visual cues as position and length. Humans are good at visually quantifying linear measures, making bar plots a strong alternative to pie or donut charts.

##Know When to Include Zero

When using bar plots, always start at 0. It is deceptive not to start at 0 because bar plots imply length is proportional to the quantity displayed. Cutting off the y-axis can make differences look bigger than they actually are.

When using position rather than length, it is not necessary to include 0 (scatterplot, dot plot, boxplot).

##Do Not Distort Quantities

Make sure your visualizations encode the correct quantities.
For example, if you are using a plot that relies on circle area, make sure the area (rather than the radius) is proportional to the quantity.

##Order by a Meaningful Value

It is easiest to visually extract information from a plot when categories are ordered by a meaningful value. The exact value on which to order will depend on your data and the message you wish to convey with your plot.
The default ordering for categories is alphabetical if the categories are strings or by factor level if factors. However, we rarely want alphabetical order.

##Show the Data

A dynamite plot - a bar graph of group averages with error bars denoting standard errors - provides almost no information about a distribution.

By showing the data, you provide viewers extra information about distributions.
Jitter is adding a small random shift to each point in order to minimize the number of overlapping points. To add jitter, use the  *geom_jitter()* geometry instead of *geom_point()*. 
*Alpha* blending is making points somewhat transparent, helping visualize the density of overlapping points. Add an alpha argument to the geometry.

```{r}
# dot plot showing the data
heights %>% ggplot(aes(sex, height)) + geom_point()

# jittered, alpha blended point plot
heights %>% ggplot(aes(sex, height)) + geom_jitter(width = 0.1, alpha = 0.2)
```

##Ease Comparisons: Use Common Axes

Ease comparisons by keeping axes the same when comparing data across multiple plots.
Align plots vertically to see horizontal changes. Align plots horizontally to see vertical changes.
Bar plots are useful for showing one number but not useful for showing distributions.

##Consider Transformations

Use transformations when warranted to ease visual interpretation.
The log transformation is useful for data with multiplicative changes. The logistic transformation is useful for fold changes in odds. The square root transformation is useful for count data.

##Ease Comparisons: Compared Visual Cues Should Be Adjacent

When two groups are to be compared, it is optimal to place them adjacent in the plot.
Use color to encode groups to be compared.
Consider using a color blind friendly palette like the one in this video.

```{r}
color_blind_friendly_cols <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

p1 <- data.frame(x = 1:8, y = 1:8, col = as.character(1:8)) %>%
    ggplot(aes(x, y, color = col)) +
    geom_point(size = 5)
p1 + scale_color_manual(values = color_blind_friendly_cols)
```

##Exercises

Using the reorder funtion inside a mutate
```{r}
library(dplyr)
library(ggplot2)
library(dslabs)
dat <- us_contagious_diseases %>% filter(year == 1967 & disease=="Measles" & count>0 & !is.na(population)) %>%
 mutate(rate = count / population * 10000 * 52 / weeks_reporting) %>%
 mutate(state = reorder(state, rate))
dat %>% ggplot(aes(state, rate)) +
 geom_bar(stat="identity") +
 coord_flip()
```

Reorder by median
```{r}
library(dplyr)
library(ggplot2)
library(dslabs)
data("murders")
murders %>% mutate(rate = total/population*100000) %>% mutate(region = reorder(region, rate,FUN = median)) %>% ggplot(aes(region,rate)) + geom_boxplot() + geom_point()
```

##Slope Charts

Consider using a slope chart or Bland-Altman plot when comparing one variable at two different time points, especially for a small number of observations.
Slope charts use angle to encode change. Use *geom_line()* to create slope charts. It is useful when comparing a small number of observations.
The *Bland-Altman plot* (Tukey mean difference plot, MA plot) graphs the difference between conditions on the y-axis and the mean between conditions on the x-axis. It is more appropriate for large numbers of observations than slope charts.

Slope Chart
```{r}
library(tidyverse)
library(dslabs)
data(gapminder)

west <- c("Western Europe", "Northern Europe", "Southern Europe", "Northern America", "Australia and New Zealand")

dat <- gapminder %>%
    filter(year %in% c(2010, 2015) & region %in% west & !is.na(life_expectancy) & population > 10^7)

dat %>%
    mutate(location = ifelse(year == 2010, 1, 2),
           location = ifelse(year == 2015 & country %in% c("United Kingdom", "Portugal"),
                             location + 0.22, location),
           hjust = ifelse(year == 2010, 1, 0)) %>%
    mutate(year = as.factor(year)) %>%
    ggplot(aes(year, life_expectancy, group = country)) +
    geom_line(aes(color = country), show.legend = FALSE) +
    geom_text(aes(x = location, label = country, hjust = hjust), show.legend = FALSE) +
    xlab("") +
    ylab("Life Expectancy") 
```

Bland-Altman plot
```{r}
library(ggrepel)
dat %>%
    mutate(year = paste0("life_expectancy_", year)) %>%
    select(country, year, life_expectancy) %>% spread(year, life_expectancy) %>%
    mutate(average = (life_expectancy_2015 + life_expectancy_2010)/2,
                difference = life_expectancy_2015 - life_expectancy_2010) %>%
    ggplot(aes(average, difference, label = country)) +
    geom_point() +
    geom_text_repel() +
    geom_abline(lty = 2) +
    xlab("Average of 2010 and 2015") +
    ylab("Difference between 2015 and 2010")
```

##