<<<<<<< HEAD
---
title: 'Section 2: Introduction to ggplot2'
author: "kevinoliva26"
date: "11/25/2020"
output: html_document
---

##ggplot2

Plots in ggplot2 consist of 3 main components:
*Data*: The dataset being summarized
*Geometry*: The type of plot (scatterplot, boxplot, barplot, histogram, qqplot, smooth density, etc.)
*Aesthetic mapping*: Variables mapped to visual cues, such as x-axis and y-axis values and color

There are additional components:
*Scale*
*Labels, Title, Legend*
*Theme/Style*

##Building a plot

You can associate a dataset x with a ggplot object with any of the 3 commands:
*ggplot(data = x)*
*ggplot(x)*
*x %>% ggplot()*

```{r}
library(tidyverse)
library(dslabs)
data(murders)

ggplot(data = murders)

murders %>% ggplot()

p <- ggplot(data = murders)
class(p)
print(p)    # this is equivalent to simply typing p
p
```
```{r}

```

##Customizing Plots

In ggplot2, graphs are created by adding layers to the ggplot object:
*DATA %>% ggplot() + LAYER_1 + LAYER_2 + ... + LAYER_N*

*aes()* uses variable names from the object component (for example, total rather than murders$total).
*geom_point()* creates a scatterplot and requires x and y aesthetic mappings.
*geom_text()* and *geom_label()* add text to a scatterplot and require x, y, and label aesthetic mappings.

To determine which aesthetic mappings are required for a geometry, read the help file for that geometry.

```{r}
library(tidyverse)
library(dslabs)
data(murders)

murders %>% ggplot() +
    geom_point(aes(x = population/10^6, y = total))
    
# add points layer to predefined ggplot object
p <- ggplot(data = murders)
p + geom_point(aes(population/10^6, total))

# add text layer to scatterplot
p + geom_point(aes(population/10^6, total)) +
    geom_text(aes(population/10^6, total, label = abb))
```

Example of aes behavior

```{r}
# no error from this call
p_test <- p + geom_text(aes(population/10^6, total, label = abb))

# error - "abb" is not a globally defined variable and cannot be found outside of aes
p_test <- p + geom_text(aes(population/10^6, total), label = abb)
```

## Tinkering

You can modify arguments to geometry functions other than aes() and the data. Additional arguments can be found in the documentation for each geometry.

Global aesthetic mappings apply to all geometries and can be defined when you initially call ggplot()

```{r}
# change the size of the points
p + geom_point(aes(population/10^6, total), size = 3) +
    geom_text(aes(population/10^6, total, label = abb))
    
# move text labels slightly to the right
p + geom_point(aes(population/10^6, total), size = 3) +
    geom_text(aes(population/10^6, total, label = abb), nudge_x = 1)
    
# simplify code by adding global aesthetic
p <- murders %>% ggplot(aes(population/10^6, total, label = abb))
p + geom_point(size = 3) +
    geom_text(nudge_x = 1.5)
    
# local aesthetics override global aesthetics
p + geom_point(size = 3) +
    geom_text(aes(x = 10, y = 800, label = "Hello there!"))
```

##Scales, Labels, and Colors

Convert the x-axis to log scale with *scale_x_continuous(trans = "log10")* or *scale_x_log10()*. Similar functions exist for the y-axis.
Add axis titles with *xlab()* and *ylab()* functions. Add a plot title with the *ggtitle()* function.
Add a color mapping that colors points by a variable by defining the col argument within aes(). To color all points the same way, define col outside of aes().
Add a line with the *geom_abline()* geometry. geom_abline() takes arguments *slope* (default = 1) and intercept (default = 0). Change the color with col or color and line type with lty.
Placing the line layer after the point layer will overlay the line on top of the points. To overlay points on the line, place the line layer before the point layer.

Code: Log-scale the x- and y-axis
```{r}
# define p
library(tidyverse)
library(dslabs)
data(murders)
p <- murders %>% ggplot(aes(population/10^6, total, label = abb))

# log base 10 scale the x-axis and y-axis
p + geom_point(size = 3) +
    geom_text(nudge_x = 0.05) +
    scale_x_continuous(trans = "log10") +
    scale_y_continuous(trans = "log10")
    
# efficient log scaling of the axes
p + geom_point(size = 3) +
    geom_text(nudge_x = 0.075) +
    scale_x_log10() +
    scale_y_log10()
```

Code: Add labels and title
```{r}
p + geom_point(size = 3) +
    geom_text(nudge_x = 0.075) +
    scale_x_log10() +
    scale_y_log10() +
    xlab("Population in millions (log scale)") +
    ylab("Total number of murders (log scale)") +
    ggtitle("US Gun Murders in 2010")
```

Code: Change color of the points
```{r}
# redefine p to be everything except the points layer
p <- murders %>%
    ggplot(aes(population/10^6, total, label = abb)) +
    geom_text(nudge_x = 0.075) +
    scale_x_log10() +
    scale_y_log10() +
    xlab("Population in millions (log scale)") +
    ylab("Total number of murders (log scale)") +
    ggtitle("US Gun Murders in 2010")
    
# make all points blue
p + geom_point(size = 3, color = "blue")

# color points by region
p + geom_point(aes(col = region), size = 3)
```

Code: Add a line with average murder rate
```{r}
# define average murder rate
r <- murders %>%
    summarize(rate = sum(total) / sum(population) * 10^6) %>%
    pull(rate)
    
# basic line with average murder rate for the country
p + geom_point(aes(col = region), size = 3) +
    geom_abline(intercept = log10(r))    # slope is default of 1
    
# change line to dashed and dark grey, line under points
p + 
    geom_abline(intercept = log10(r), lty = 2, color = "darkgrey") +
    geom_point(aes(col = region), size = 3)
```

Code: Change legend title
```{r}
p <- p + scale_color_discrete(name = "Region")    # capitalize legend title
p
```

##Add-on Packages

The style of a ggplot graph can be changed using the *theme()* function.
The ggthemes package adds additional themes.
The ggrepel package includes a geometry that repels text labels, ensuring they do not overlap with each other: *geom_text_repel()*.

```{r}
# theme used for graphs in the textbook and course
library(dslabs)
ds_theme_set()

# themes from ggthemes
library(ggthemes)
p + theme_economist()    # style of the Economist magazine
p + theme_fivethirtyeight()    # style of the FiveThirtyEight website
```

Putting it all toghether
```{r}
# load libraries
library(tidyverse)
library(ggrepel)
library(ggthemes)
library(dslabs)
data(murders)

# define the intercept
r <- murders %>%
    summarize(rate = sum(total) / sum(population) * 10^6) %>%
    .$rate #.$rate es lo mismo que pull(rate)
    
# make the plot, combining all elements
murders %>%
    ggplot(aes(population/10^6, total, label = abb)) +
    geom_abline(intercept = log10(r), lty = 2, color = "darkgrey") +
    geom_point(aes(col = region), size = 3) +
    geom_text_repel() +
    scale_x_log10() +
    scale_y_log10() +
    xlab("Population in millions (log scale)") +
    ylab("Total number of murders (log scale)") +
    ggtitle("US Gun Murders in 2010") +
    scale_color_discrete(name = "Region") +
    theme_economist()
```

##Other examples

*geom_histogram()* creates a histogram. Use the binwidth argument to change the width of bins, the fill argument to change the bar fill color, and the col argument to change bar outline color.
*geom_density()* creates smooth density plots. Change the fill color of the plot with the fill argument.
*geom_qq()* creates a quantile-quantile plot. This geometry requires the sample argument. By default, the data are compared to a standard normal distribution with a mean of 0 and standard deviation of 1. 
Plots can be arranged adjacent to each other using the *grid.arrange()* function from the gridExtra package.

Histograms
```{r}
# load heights data
library(tidyverse)
library(dslabs)
data(heights)

# define p
p <- heights %>%
    filter(sex == "Male") %>%
    ggplot(aes(x = height))
    
# basic histograms
p + geom_histogram()
p + geom_histogram(binwidth = 1)

# histogram with blue fill, black outline, labels and title
p + geom_histogram(binwidth = 1, fill = "blue", col = "black") +
    xlab("Male heights in inches") +
    ggtitle("Histogram")
```

Smooth Density
```{r}
p + geom_density()
p + geom_density(fill = "blue")
```

Quantile-quantile plots in ggplot2
```{r}
# basic QQ-plot
p <- heights %>% filter(sex == "Male") %>%
    ggplot(aes(sample = height))
p + geom_qq()

# QQ-plot against a normal distribution with same mean/sd as data
params <- heights %>%
    filter(sex == "Male") %>%
    summarize(mean = mean(height), sd = sd(height))
    p + geom_qq(dparams = params) +
    geom_abline()
    
# QQ-plot of scaled data against the standard normal distribution
heights %>%
    ggplot(aes(sample = scale(height))) +
    geom_qq() +
    geom_abline()
```

Grids of plots with the gridExtra package
```{r}
# define plots p1, p2, p3
p <- heights %>% filter(sex == "Male") %>% ggplot(aes(x = height))
p1 <- p + geom_histogram(binwidth = 1, fill = "blue", col = "black")
p2 <- p + geom_histogram(binwidth = 2, fill = "blue", col = "black")
p3 <- p + geom_histogram(binwidth = 3, fill = "blue", col = "black")

# arrange plots next to each other in 1 row, 3 columns
library(gridExtra)
grid.arrange(p1, p2, p3, ncol = 3)
```
=======
---
title: 'Section 2: Introduction to ggplot2'
author: "kevinoliva26"
date: "11/25/2020"
output: html_document
---

##ggplot2

Plots in ggplot2 consist of 3 main components:
*Data*: The dataset being summarized
*Geometry*: The type of plot (scatterplot, boxplot, barplot, histogram, qqplot, smooth density, etc.)
*Aesthetic mapping*: Variables mapped to visual cues, such as x-axis and y-axis values and color

There are additional components:
*Scale*
*Labels, Title, Legend*
*Theme/Style*

##Building a plot

You can associate a dataset x with a ggplot object with any of the 3 commands:
*ggplot(data = x)*
*ggplot(x)*
*x %>% ggplot()*

```{r}
library(tidyverse)
library(dslabs)
data(murders)

ggplot(data = murders)

murders %>% ggplot()

p <- ggplot(data = murders)
class(p)
print(p)    # this is equivalent to simply typing p
p
```
```{r}

```

##Customizing Plots

In ggplot2, graphs are created by adding layers to the ggplot object:
*DATA %>% ggplot() + LAYER_1 + LAYER_2 + ... + LAYER_N*

*aes()* uses variable names from the object component (for example, total rather than murders$total).
*geom_point()* creates a scatterplot and requires x and y aesthetic mappings.
*geom_text()* and *geom_label()* add text to a scatterplot and require x, y, and label aesthetic mappings.

To determine which aesthetic mappings are required for a geometry, read the help file for that geometry.

```{r}
library(tidyverse)
library(dslabs)
data(murders)

murders %>% ggplot() +
    geom_point(aes(x = population/10^6, y = total))
    
# add points layer to predefined ggplot object
p <- ggplot(data = murders)
p + geom_point(aes(population/10^6, total))

# add text layer to scatterplot
p + geom_point(aes(population/10^6, total)) +
    geom_text(aes(population/10^6, total, label = abb))
```

Example of aes behavior

```{r}
# no error from this call
p_test <- p + geom_text(aes(population/10^6, total, label = abb))

# error - "abb" is not a globally defined variable and cannot be found outside of aes
p_test <- p + geom_text(aes(population/10^6, total), label = abb)
```

## Tinkering

You can modify arguments to geometry functions other than aes() and the data. Additional arguments can be found in the documentation for each geometry.

Global aesthetic mappings apply to all geometries and can be defined when you initially call ggplot()

```{r}
# change the size of the points
p + geom_point(aes(population/10^6, total), size = 3) +
    geom_text(aes(population/10^6, total, label = abb))
    
# move text labels slightly to the right
p + geom_point(aes(population/10^6, total), size = 3) +
    geom_text(aes(population/10^6, total, label = abb), nudge_x = 1)
    
# simplify code by adding global aesthetic
p <- murders %>% ggplot(aes(population/10^6, total, label = abb))
p + geom_point(size = 3) +
    geom_text(nudge_x = 1.5)
    
# local aesthetics override global aesthetics
p + geom_point(size = 3) +
    geom_text(aes(x = 10, y = 800, label = "Hello there!"))
```

##Scales, Labels, and Colors

Convert the x-axis to log scale with *scale_x_continuous(trans = "log10")* or *scale_x_log10()*. Similar functions exist for the y-axis.
Add axis titles with *xlab()* and *ylab()* functions. Add a plot title with the *ggtitle()* function.
Add a color mapping that colors points by a variable by defining the col argument within aes(). To color all points the same way, define col outside of aes().
Add a line with the *geom_abline()* geometry. geom_abline() takes arguments *slope* (default = 1) and intercept (default = 0). Change the color with col or color and line type with lty.
Placing the line layer after the point layer will overlay the line on top of the points. To overlay points on the line, place the line layer before the point layer.

Code: Log-scale the x- and y-axis
```{r}
# define p
library(tidyverse)
library(dslabs)
data(murders)
p <- murders %>% ggplot(aes(population/10^6, total, label = abb))

# log base 10 scale the x-axis and y-axis
p + geom_point(size = 3) +
    geom_text(nudge_x = 0.05) +
    scale_x_continuous(trans = "log10") +
    scale_y_continuous(trans = "log10")
    
# efficient log scaling of the axes
p + geom_point(size = 3) +
    geom_text(nudge_x = 0.075) +
    scale_x_log10() +
    scale_y_log10()
```

Code: Add labels and title
```{r}
p + geom_point(size = 3) +
    geom_text(nudge_x = 0.075) +
    scale_x_log10() +
    scale_y_log10() +
    xlab("Population in millions (log scale)") +
    ylab("Total number of murders (log scale)") +
    ggtitle("US Gun Murders in 2010")
```

Code: Change color of the points
```{r}
# redefine p to be everything except the points layer
p <- murders %>%
    ggplot(aes(population/10^6, total, label = abb)) +
    geom_text(nudge_x = 0.075) +
    scale_x_log10() +
    scale_y_log10() +
    xlab("Population in millions (log scale)") +
    ylab("Total number of murders (log scale)") +
    ggtitle("US Gun Murders in 2010")
    
# make all points blue
p + geom_point(size = 3, color = "blue")

# color points by region
p + geom_point(aes(col = region), size = 3)
```

Code: Add a line with average murder rate
```{r}
# define average murder rate
r <- murders %>%
    summarize(rate = sum(total) / sum(population) * 10^6) %>%
    pull(rate)
    
# basic line with average murder rate for the country
p + geom_point(aes(col = region), size = 3) +
    geom_abline(intercept = log10(r))    # slope is default of 1
    
# change line to dashed and dark grey, line under points
p + 
    geom_abline(intercept = log10(r), lty = 2, color = "darkgrey") +
    geom_point(aes(col = region), size = 3)
```

Code: Change legend title
```{r}
p <- p + scale_color_discrete(name = "Region")    # capitalize legend title
p
```

##Add-on Packages

The style of a ggplot graph can be changed using the *theme()* function.
The ggthemes package adds additional themes.
The ggrepel package includes a geometry that repels text labels, ensuring they do not overlap with each other: *geom_text_repel()*.

```{r}
# theme used for graphs in the textbook and course
library(dslabs)
ds_theme_set()

# themes from ggthemes
library(ggthemes)
p + theme_economist()    # style of the Economist magazine
p + theme_fivethirtyeight()    # style of the FiveThirtyEight website
```

Putting it all toghether
```{r}
# load libraries
library(tidyverse)
library(ggrepel)
library(ggthemes)
library(dslabs)
data(murders)

# define the intercept
r <- murders %>%
    summarize(rate = sum(total) / sum(population) * 10^6) %>%
    .$rate #.$rate es lo mismo que pull(rate)
    
# make the plot, combining all elements
murders %>%
    ggplot(aes(population/10^6, total, label = abb)) +
    geom_abline(intercept = log10(r), lty = 2, color = "darkgrey") +
    geom_point(aes(col = region), size = 3) +
    geom_text_repel() +
    scale_x_log10() +
    scale_y_log10() +
    xlab("Population in millions (log scale)") +
    ylab("Total number of murders (log scale)") +
    ggtitle("US Gun Murders in 2010") +
    scale_color_discrete(name = "Region") +
    theme_economist()
```

##Other examples

*geom_histogram()* creates a histogram. Use the binwidth argument to change the width of bins, the fill argument to change the bar fill color, and the col argument to change bar outline color.
*geom_density()* creates smooth density plots. Change the fill color of the plot with the fill argument.
*geom_qq()* creates a quantile-quantile plot. This geometry requires the sample argument. By default, the data are compared to a standard normal distribution with a mean of 0 and standard deviation of 1. 
Plots can be arranged adjacent to each other using the *grid.arrange()* function from the gridExtra package.

Histograms
```{r}
# load heights data
library(tidyverse)
library(dslabs)
data(heights)

# define p
p <- heights %>%
    filter(sex == "Male") %>%
    ggplot(aes(x = height))
    
# basic histograms
p + geom_histogram()
p + geom_histogram(binwidth = 1)

# histogram with blue fill, black outline, labels and title
p + geom_histogram(binwidth = 1, fill = "blue", col = "black") +
    xlab("Male heights in inches") +
    ggtitle("Histogram")
```

Smooth Density
```{r}
p + geom_density()
p + geom_density(fill = "blue")
```

Quantile-quantile plots in ggplot2
```{r}
# basic QQ-plot
p <- heights %>% filter(sex == "Male") %>%
    ggplot(aes(sample = height))
p + geom_qq()

# QQ-plot against a normal distribution with same mean/sd as data
params <- heights %>%
    filter(sex == "Male") %>%
    summarize(mean = mean(height), sd = sd(height))
    p + geom_qq(dparams = params) +
    geom_abline()
    
# QQ-plot of scaled data against the standard normal distribution
heights %>%
    ggplot(aes(sample = scale(height))) +
    geom_qq() +
    geom_abline()
```

Grids of plots with the gridExtra package
```{r}
# define plots p1, p2, p3
p <- heights %>% filter(sex == "Male") %>% ggplot(aes(x = height))
p1 <- p + geom_histogram(binwidth = 1, fill = "blue", col = "black")
p2 <- p + geom_histogram(binwidth = 2, fill = "blue", col = "black")
p3 <- p + geom_histogram(binwidth = 3, fill = "blue", col = "black")

# arrange plots next to each other in 1 row, 3 columns
library(gridExtra)
grid.arrange(p1, p2, p3, ncol = 3)
```
>>>>>>> c1ea3ff55daacaf93b5a58b750522a886d7a3442

