---
title: 'Section 1: Introduction to Data Visualization and Distributions'
author: "kevinoliva26"
date: "11/23/2020"
output: html_document
---

##1.1 Introduction to Data Visualization

Categorical data are variables that are defined by a small number of groups.
  Ordinal categorical data have an inherent order to the categories (mild/medium/hot, for example).
  Non-ordinal categorical data have no order to the categories.
Numerical data take a variety of numeric values.
  Continuous variables can take any value.
  Discrete variables are limited to sets of specific values.

For categorical data we can construct this distribution by simply computing the frequency of each unique value. This can be done with the function *table*

```{r}
library(dslabs)
data(heights)
x <- heights$height
head(table(x))
```

  
##1.2 Introduction to Distributions


```{r}
# load the dataset
library(dslabs)
data(heights)

# make a table of category proportions
prop.table(table(heights$sex))  
```

Every continuous distribution has a cumulative distribution function (CDF). The CDF defines the proportion of the data below a given value  a  for all values of  a :

```{r}
data(heights)
my_data <- heights$height
a <- seq(min(my_data), max(my_data), length = 100)    # define range of values spanning the dataset
cdf_function <- function(x) {    # computes prob. for a single value
    mean(my_data <= x)
}
cdf_values <- sapply(a, cdf_function)
plot(a, cdf_values)
```

##Smooth Density Plots

Smooth density plots can be thought of as histograms where the bin width is extremely or infinitely small. The smoothing function makes estimates of the true continuous trend of the data given the available sample of data points.

The y-axis is scaled so that the area under the density curve sums to 1.

##Normal Distribution

Standard units describe how many standard deviations a value is away from the mean. The z-score, or number of standard deviations an observation  x  is away from the mean  μ :
Z=x−μσ
Compute standard units with the *scale()* function.

```{r}
# define x as vector of male heights
library(tidyverse)
library(dslabs)
data(heights)
index <- heights$sex=="Male"
x <- heights$height[index]

# calculate the mean and standard deviation manually
average <- sum(x)/length(x)
SD <- sqrt(sum((x - average)^2)/length(x))

# built-in mean and sd functions - note that the audio and printed values disagree
average <- mean(x)
SD <- sd(x)
c(average = average, SD = SD)

# calculate standard units
z <- scale(x)

# calculate proportion of values within 2 SD of mean
mean(abs(z) < 2)
```

The normal distribution is associated with the 68-95-99.7 rule. This rule describes the probability of observing events within a certain number of standard deviations of the mean. 

##The normal CDF and pnorm

*pnorm(a, avg, s)* gives the value of the cumulative distribution function  F(a)  for the normal distribution defined by average avg and standard deviation s.

We say that a random quantity is normally distributed with average avg and standard deviation s if the approximation pnorm(a, avg, s) holds for all values of a.

If we treat the height data as discrete rather than categorical, we see that the data are not very useful because integer values are more common than expected due to rounding. This is called *discretization*.

```{r}
library(tidyverse)
library(dslabs)
data(heights)
x <- heights %>% filter(sex=="Male") %>% pull(height)

#We can estimate the probability that a male is taller than 70.5 inches with:
1 - pnorm(70.5, mean(x), sd(x))

# plot distribution of exact heights in data
plot(prop.table(table(x)), xlab = "a = Height in inches", ylab = "Pr(x = a)")

# probabilities in actual data over length 1 ranges containing an integer
mean(x <= 68.5) - mean(x <= 67.5)
mean(x <= 69.5) - mean(x <= 68.5)
mean(x <= 70.5) - mean(x <= 69.5)

# probabilities in normal approximation match well
pnorm(68.5, mean(x), sd(x)) - pnorm(67.5, mean(x), sd(x))
pnorm(69.5, mean(x), sd(x)) - pnorm(68.5, mean(x), sd(x))
pnorm(70.5, mean(x), sd(x)) - pnorm(69.5, mean(x), sd(x))

# probabilities in actual data over other ranges don't match normal approx as well
mean(x <= 70.9) - mean(x <= 70.1)
pnorm(70.9, mean(x), sd(x)) - pnorm(70.1, mean(x), sd(x))
```

##1.3 Quantiles, Percentiles, and Boxplots

```{r}
data <- heights$height
p <- seq(0.01, 0.99, 0.01)
quantile(data, p)
```

Use *summary()* on the heights$height variable to find the quartiles:

```{r}
summary(heights$height)
p <- seq(0.01, 0.99, 0.01)
percentiles <- quantile(heights$height, p)
percentiles[names(percentiles) == "25%"]
percentiles[names(percentiles) == "75%"]
```

##Finding quantiles with qnorm

The *qnorm()* function gives the theoretical value of a quantile with probability p of observing a value equal to or less than that quantile value given a normal distribution with mean mu and standard deviation sigma:

*qnorm(p, mu, sigma)*

By default, mu=0 and sigma=1. Therefore, calling qnorm() with no arguments gives quantiles for the standard normal distribution.

The *pnorm()* function gives the probability that a value from a standard normal distribution will be less than or equal to a z-score value z. Consider:

pnorm(-1.96)  ≈0.025 
The result of pnorm() is the quantile. Note that:

qnorm(0.025)  ≈−1.96
qnorm() and pnorm() are inverse functions:

pnorm(qnorm(0.025)) = 0.025

##Quantile-Quantile Plots

Quantile-quantile plots, or QQ-plots, are used to check whether distributions are well-approximated by a normal distribution.

Calculate sample quantiles (observed quantiles) using the *quantile()* function.
Calculate theoretical quantiles with the *qnorm()* function. qnorm() 
```{r}
# define x and z
library(tidyverse)
library(dslabs)
data(heights)
index <- heights$sex=="Male"
x <- heights$height[index]
z <- scale(x)

# proportion of data below 69.5
mean(x <= 69.5)

# calculate observed and theoretical quantiles
p <- seq(0.05, 0.95, 0.05)
observed_quantiles <- quantile(x, p)
theoretical_quantiles <- qnorm(p, mean = mean(x), sd = sd(x))

# make QQ-plot
plot(theoretical_quantiles, observed_quantiles)
abline(0,1)

# make QQ-plot with scaled values
observed_quantiles <- quantile(z, p)
theoretical_quantiles <- qnorm(p)
plot(theoretical_quantiles, observed_quantiles)
abline(0,1)
```



